Delta Smelt Enclosure Study 2019 Hybrid Paper Diet Analysis

### 2023-01-26 smelt_2019_allseasons/Zoop 2019.Rmd
added multivariate stats

---
librarytitle: "Zoop Plotting"
output: html_document
editor_options: 
  chunk_output_type: console
---
# load packages
```{r}
library(tidyverse)
library(lubridate)
library(Polychrome)
```

# load & prep summer/fall data
```{r}
zoop <- read_csv("smelt_2019_allseasons/data_raw/ZoopData_SummerFall_2019.csv")

@@

#bring in tow data
aug <- read_csv("smelt_2019_allseasons/data_raw/Field_Environmental_Data_Aug_qaqc'ed_11.04.20.csv")
oct <- read_csv("smelt_2019_allseasons/data_raw/Field_Environmental_Data_Oct_qaqc'ed_12.11.20.csv")
aug$Date <- mdy(aug$Date)
oct$Date <- mdy(oct$Date)

@@

winter <- read_csv("smelt_2019_allseasons/data_clean/zoop_abundance.csv")

@@

>>>>

---
librarytitle: "Zoop Plotting"
output: html_document
editor_options: 
  chunk_output_type: console
---
# load packages
```{r}
library(tidyverse)
library(lubridate)
library(Polychrome)
library(here)
```

# load & prep summer/fall data
```{r}

#zoop <- read_csv("smelt_2019_allseasons/data_raw/ZoopData_SummerFall_2019.csv")
zoop <- read_csv("data_raw/ZoopData_SummerFall_2019.csv")

@@

#bring in tow data
#aug <- read_csv("smelt_2019_allseasons/data_raw/Field_Environmental_Data_Aug_qaqc'ed_11.04.20.csv")
#oct <- read_csv("smelt_2019_allseasons/data_raw/Field_Environmental_Data_Oct_qaqc'ed_12.11.20.csv")
aug <- read_csv("data_raw/Field_Environmental_Data_Aug_qaqc'ed_11.04.20.csv")
oct <- read_csv("data_raw/Field_Environmental_Data_Oct_qaqc'ed_12.11.20.csv")

aug$Date <- mdy(aug$Date)
oct$Date <- mdy(oct$Date)

@@

#winter <- read_csv("smelt_2019_allseasons/data_clean/zoop_abundance.csv")
winter <- read_csv("data_clean/zoop_abundance.csv")

@@

## Multivariate statistics

```{r}

library(vegan)
source("../PlotNMDS.R")

zoops = mutate(caught.19, SampleID = paste(Location, Date)) %>%
  pivot_wider(id_cols = c(SampleID, Location, Date, MD, Season), names_from = Taxa, values_from = ind_m3, values_fn = sum, values_fill = 0) %>%
  mutate(datatype = "Zoops")

#first let's look at Rio Vista across seasons
RVzoops = filter(zoops, Location == "RV")

#raw abundance matrix
RVzoopmat = as.matrix(RVzoops[,6:12])

#relative abundance matrix
RVzoopmat2 = RVzoopmat/rowSums(RVzoopmat)

RVa = adonis2(RVzoopmat~Season, data = RVzoops)
RVa
#

RVb = adonis2(RVzoopmat2~Season, data = RVzoops)
RVb

```
pretty big differences between seasons, R2 of 54% on the absolute abundance, 64% on relative abundance

THough I may need to update this with some sort of variable to account for day of year. Not sure.

Now I'll look for changes by location

```{r}
Fallzoops = filter(zoops, Season == "Fall")

#raw abundance matrix
Fallzoopmat = as.matrix(Fallzoops[,6:12])

#relative abundance matrix
Fallzoopmat2 = Fallzoopmat/rowSums(Fallzoopmat)

Falla = adonis2(Fallzoopmat~Location, data = Fallzoops)
Falla
#

Fallb = adonis2(Fallzoopmat2~Location, data = Fallzoops)
Fallb
#

```

Wow! Even larger differences in relative and absolute abundance by region

I'm going to try an NMDS just for funzies

```{r}

#raw abundance matrix
zoopmat = as.matrix(zoops[,6:12])

#relative abundance matrix
zoopmat2 = zoopmat/rowSums(zoopmat)

N1 = metaMDS(zoopmat2)
zoops = mutate(zoops, Season  = as.factor(Season), Location = as.factor(Location))
PlotNMDS(N1, zoops, group = "Season")
PlotNMDS(N1, zoops, group = "Location")

RVN1 = metaMDS(RVzoopmat2)
RVzoops = mutate(RVzoops, Season  = as.factor(Season), Location = as.factor(Location))
PlotNMDS(RVN1, RVzoops, group = "Season")
PlotNMDS(RVN1, RVzoops, group = "Location")

```

## Diet data comparison

```{r}
#diet2 <- read_csv("smelt_2019_allseasons/data_raw/DSM Cage Diets 2019-2020.csv") #new data, not edited at all
diet2 <- read_csv("data_raw/DSM Cage Diets 2019-2020.csv") #new data, not edited at all

diet2$Date <- mdy(diet2$Date)

#format 
#rename columns
diet2 <- diet2 %>% rename(Cage.ID = "Cage ID", Contents.Weight = "Total Contents Weight", Prey.Taxa = "Prey Taxa", LH.Stage = "LH Stage")

#add taxa column (choose taxa based on what is present in the dataframe)
diet2$Taxa <- as.character(NA)

str(diet2)

diet2[diet2$Prey.Taxa %in% c("Bosmina sp.", "Chydorus sp.", "Daphnia sp.", "Ceriodaphnia sp.", "Chydoridae", "Crangonyx sp.", "Scapholeberis sp.", "Simocephalus sp.", "Eurycercus sp.", "Leydigia sp.", "Cladocera", "Sida sp.", "Alona sp.", "Diaphanosoma sp."),]$Taxa <- "Cladocera"

diet2[diet2$Prey.Taxa %in% c("Americorophium sp.", "Americorophium spinicorne", "Eogammarus confervicolus", "Gammaridea", "Hyallela azteca", "Crangonyx sp.", "Corophiidae", "Amphipoda"),]$Taxa <- "Amphipod"

diet2[diet2$Prey.Taxa %in% c("Pseudodiaptomus forbesi", "Calanoida", "Eurytemora affinis", "Diaptomidae", "Sinocalanus doerrii", "Osphranticum labronectum"),]$Taxa <- "Calanoid"

diet2[diet2$Prey.Taxa %in% c("Acanthocyclops sp.", "Macrocyclops sp.", "Diacyclops thomasi", "Eucyclops sp.", "Cyclopoida", "Halicyclops sp.", "Mesocyclops sp.", "Tropocyclops sp.", "Tropocyclops sp."),]$Taxa <- "Cyclopoid"

diet2[diet2$Prey.Taxa %in% c("Chironomidae", "Diptera"),]$Taxa <- "Diptera"

diet2[diet2$Prey.Taxa %in% c("empty", "EMPTY"),]$Taxa <- "empty"

diet2[diet2$Prey.Taxa %in% c("Harpacticoida"),]$Taxa <- "Harpacticoid"

diet2[diet2$Prey.Taxa %in% c("Ostracoda"),]$Taxa <- "Ostracod"

diet2[diet2$Prey.Taxa %in% c("Fish", "Acari", "Insecta", "Bivalvia", "Culicidae", "egg unidentified", "Corixidae", "Hemiptera", "plant seed", "plant matter", "Unidentified", "Mysidacea", "Cumacea"),]$Taxa <- "Other"

new_DF <- diet2[is.na(diet2$Taxa),] #update code above with new prey.taxa that weren't found in previous diets
#there are two "copepoda" entries with no count or location...don't seem that important so I am going to just go ahead and remove them
diet2[diet2$Taxa %in% c("empty"),]$Count <- "0"
diet2 <- diet2[!is.na(diet2$Count),]

#Tag column currently has rows with just the Tag and others with location, date, and tag
#first break those two kinds of columns apart
weird.tag <- diet2[is.na(diet2$Location),]
norm.tag <- diet2[!is.na(diet2$Location),]

#then for break out the Tag ID into Location and tag (remove date)
weird.tag <- weird.tag %>% separate(Tag, c("Location", "X.date", "Tag"))

#remove new date column
weird.tag <- subset(weird.tag, select = -6)

#add mesh types for prototype experiment
#create new column for mesh types
Cage_type <- as.data.frame(matrix(0, ncol = 2, nrow = 6))
colnames(Cage_type) <- c("Cage.ID", "Mesh")

Cage_type$Cage.ID <- c("A", "B", "C", "D", "E", "F")
Cage_type$Mesh <- c("small", "large", "wrap", "wrap", "small", "large")

norm.tag2 <- merge(norm.tag, Cage_type, by = "Cage.ID", all.x = TRUE)

#add mesh types (NA) for summer/fall experiements
weird.tag$Mesh <- NA

#merge the two
diet <- rbind(weird.tag, norm.tag2)

#change location titles 
diet = mutate(diet, Location2 = case_when(Location %in% c("RV", "RIVERS") ~ "RV", 
                                       Location == "Sac DWSC" ~ "DWSC",
                                       Location == "SM" ~ "SM",
                                       Location == "YB" ~ "YB"),
              Month = month(Date),
              Season = case_when(Month %in% c(2,3,5) ~ "Winter",
                                 Month == 8 ~ "Summer",
                                 Month == 11 ~ "Fall"),
              Location = Location2,
              datatype = "diet")

#OMG DON"T PUT COMMENTS IN THE COUNT COLUMN
diet$Count[which(diet$Count == "1 maybe 2")] = "1"
diet$Count = as.numeric(diet$Count)

#pivot_wider
dietwide = mutate(diet, SampleID = paste(Location, Tag)) %>%
  pivot_wider(id_cols = c(SampleID, Location, Date, Season, datatype), names_from = Taxa, values_from = Count, values_fn = sum, values_fill = 0)

```

Now put the zoop data and diet data together

```{r}

zoopdiet = bind_rows(zoops, dietwide) %>%
  mutate(Amphipod = case_when(is.na(Amphipod) ~0, TRUE ~ Amphipod),
         Diptera = case_when(is.na(Diptera) ~0, TRUE ~ Diptera)) %>%
  select(-empty, -Microzoo)

rowSums(zoopdiet[,c(6:11, 13, 14)])

#remove empty rows
zoopdiet = zoopdiet[which(rowSums(zoopdiet[,c(6:11, 13, 14)]) !=0),] %>%
  mutate(datatype = as.factor(datatype), Season = as.factor(Season), Location = as.factor(Location))

#Convert to reletive abundance
zoopdietmat = as.matrix(zoopdiet[,c(6:11, 13, 14)])
zoopdietmat2 = zoopdietmat/rowSums(zoopdietmat)

#look at just Rio Vista
RVdietzoops = filter(zoopdiet, Location == "RV")

#Convert to reletive abundance
RVzoopdietmat = as.matrix(RVdietzoops[,c(6:11, 13, 14)])
RVzoopdietmat2 = RVzoopdietmat/rowSums(RVzoopdietmat)

RVda = adonis2(RVzoopdietmat2~ datatype + Season, data = RVdietzoops)
RVda

zdM = metaMDS(zoopdietmat2)

PlotNMDS(zdM, data = zoopdiet, group = "datatype")

PlotNMDS(zdM, data = zoopdiet, group = "Location")

RVM = metaMDS(RVzoopdietmat2, trymax = 100)

PlotNMDS(RVM, data = RVdietzoops, group = "Season")
```

Basic plot of zoops versus diet

```{r}
zoopdietlong = pivot_longer(zoopdiet, cols = c(Ostracod, Cyclopoid, Calanoid, Cladocera, Harpacticoid, Other, Amphipod, Diptera),
                            names_to = "Taxa", values_to= "count")

ggplot(zoopdietlong, aes(x = datatype, y = count, fill = Taxa))+
  geom_col(position = "fill")+
  facet_grid(Location~Season)

```
